Resultants and subresultants of p-adic polynomials

### 2014-12-12 resultants/resultants.tex

\usepackage{amsmath,amssymb}
\usepackage{amsrefs}
\usepackage[usenames,dvipsnames]{color}
\usepackage{stmaryrd}
\usepackage{enumerate}
%\usepackage[algoruled,vlined,english,linesnumbered]{algorithm2e}
\usepackage[pdfpagelabels,colorlinks=true,citecolor=blue]{hyperref}
\usepackage{comment}
\usepackage{tikz}
\usepackage{bbm}
\renewcommand*{\theHsection}{\thesection}

@@

\newcommand{\ring}{\mathfrak A}
\newcommand{\fracring}{\mathfrak K}
\renewcommand{\prec}{\text{\rm prec}}
\renewcommand{\val}{\text{\rm val}}
\newcommand{\id}{\text{\rm id}}
\newcommand{\Res}{\text{\rm Res}}
\newcommand{\lc}{\text{\rm lc}}

@@

\title{Resultants and subresultants\\of p-adic polynomials}

@@

\begin{abstract}
\end{abstract}

@@

 \vspace{1mm}
 \noindent
 {\bf Keywords:} 
\medskip

@@

The strating point of this paper is the willing to desing a reasonable 
algorithm to compute \textsc{gcd}s and B\'ezout coefficients of 
polynomials with coefficients in $\Zp$ (the ring of $p$-adic integers). 
Indeed, easy experimentations show that all standard algorithms for this 
(\emph{e.g.} extended Euclide's algorithm) are not satisfying because of 
stability.

Stability issues occur in this context because, just like real numbers, 
$p$-adic numbers are by nature infinite and thus need to be truncated in 
order to fit in the memory of a computer. The level of truncation is 
called the \emph{precision}. Usual software implementing $p$-adics (as 
\textsc{magma}, \textsc{sage}) generally tracks the precision as follows:
an individual precision is attached to any $p$-adic variable and this
precision is updated after each basic arithmetic operation (addition
and multiplication). This way of tracking precision can be seen as the
analogue of the arithmetic intervals in the real setting.

Figure \ref{fig:} illustrates the unstability of Euclide's algorithm 
which it is called on random inputs which are monic $2$-adic polynomials 
of fixed degree. Let us explain how this table should be read by 
commenting on its last line. It indicates that (extended) Euclide's 
algorithm outputs the B\'ezout coefficients of two monic $2$-adic 
polynomials of degree $500$ with an average loss of $...$ significant 
digits by coefficient whereas a stable algorithm should only loose $...$ 
digits in average. This ``theoretical'' loss is computed as the double 
of the valuation of the resultant. Indeed Cramer-like formulae imply 
that B\'ezout coefficients can be computed by performing a unique 
division by the resultant, inducing then only the aforementioned loss 
of precision. Examining the table a bit more, we furthermore observe 
that the ``practical'' loss of precision due to Euclide's algorithm 
seems to grow linearly with respect to the degree whereas the 
``theoretical'' loss seems to be independant on the degree. In other
words, the unstability of Euclide's algorithm is becoming more and
more critical when the degree of the input increases.

@@

[...]

@@

@@

bla

@@

\noindent
\textbf{Definitions.}
Let $\A$ be a \emph{discrete valuation ring}. We recall that it is a 
domain equipped with a map $\val : \A \to \Z \cup \{+\infty\}$ --- the 
so-called \emph{valuation} --- satisfying the four axioms: (1)~$\val(x) = 
+\infty$ iff $x = 0$, (2)~$\val(xy) = \val(x) + \val(y)$ and (3)~$\val(x+y) \geq 
\min(\val(x), \val(y))$, (4)~any element of valuation $0$ is invertible.
We also recall that a discrete valuation ring admits a unique maximal 
ideal $\m$, consisting of elements of positive valuation. 
Furthermore, this ideal is principal and generated by any element of 
minimal valuation. Such an element is called a \emph{uniformizer}.
Let us fix one of them and denote it by $\pi$. The \emph{residue
field} of $W$ is the quotient $W/\m = W/\pi W$ and we shall denote
it by $k$.

The valuation defines a distance $d$ on $\A$ by letting $d(x,y) = 
e^{-\val(x-y)}$ for all $x, y \in \A$. We say that $\A$ is \emph{complete}
if it is complete with respect to the previous distance. Assuming that
$\A$ is complete, any element $x \in \A$ can be written uniquely as a 
convergent series:
\begin{equation}
\label{eq:expandCDVR}
x = \sum_{i=0}^\infty x_i \pi^i
\end{equation}
where the $x_i$'s lie in a fixed set of representatives of classes
modulo $\pi$. Therefore, as an additive group, $W$ is isomorphic to 
the set of sequences $\N \to k$. On the contrary, the multiplicative
structure may vary.

Let $K$ denote the fraction field of $\A$. The valuation $v$ extends 
uniquely to $K$ by letting $\val(\frac x y) = \val(x) - \val(y)$. 
Moreover, it follows from the axiom (4) that $K$ is obtain from $\A$ by 
inverting $\pi$. Thus, any element of $K$ can be uniquely written as an 
infinite sum:
\begin{equation}
\label{eq:expandCDVF}
x = \sum_{i=v}^\infty x_i \pi^i
\end{equation}
where $v$ is some relative integer and the $x_i$'s are as above. The
valuation of $x$ can be easily read on this writing: it is the smallest 
integer $v$ such that $x_i \not\equiv 0 \pmod \pi$.

@@

\noindent
\textbf{Examples.}
The two main examples of complete discrete valuation rings are:
\begin{enumerate}[(1)]
\vspace{-0.25cm}
\item the ring of $p$-adic integers $\Zp$: the valuation on it is
the usual $p$-adic valuation, a uniformizer is $p$, its residue
field is $\Fp = \Z/p\Z$ and Eq.~\eqref{eq:expandCDVR} corresponds
to standard expansion in $p$-basis of a $p$-adic number; the fraction 
field of $\Zp$ is the so-called field of $p$-adic numbers and is usually 
denoted by $\Qp$;
\vspace{-0.25cm}
\item the ring of formal series $k[[x]]$ where $k$ is a field: the
valuation on it is the usual valuation of series, a uniformizer is
$x$, the residue field is $k$ and Eq.~\eqref{eq:expandCDVR} is the
usual writing of formal series; the fraction field of $k[[x]]$ is
the field of Laurent series $k((x))$.
\end{enumerate}
\vspace{-0.2cm}

\noindent
\textbf{Precision.}
The memory of a computer being necessarily finite, it cannot be possible 
to find a way to represent exhaustively all elements of $\A$. 
Very often, mimicing what we do for real numbers, we choose to truncate 
the expansion \eqref{eq:expandCDVR} at some finite level. Concretely,
this means that we work with approximations of elements of $\A$ of the
form:
$$\sum_{i=0}^{N-1} x_i \pi^i + O(\pi^N)$$
where $N$ is some integer called the \emph{precision} of $x$ and 
the term $O(\pi^N)$ means that the $x_i$'s with $i \geq N$ are not 
specified. Hence, the notation above does not represent a unique
element of $\A$ but all elements lying in the ball having center
$\sum_{i=0}^{N-1} x_i \pi^i$ and radius $e^{-N v(\pi)}$. In other
words, on a computer, we do not work with elements of $\A$ but with
balls in $\A$ which are much more tractable.

\subsection{Subresultants}

A first issue when dealing with numerical computations of \textsc{gcd}s 
of polynomials over $\A$ is that the $\gcd$ function is not continuous: 
it takes the value $1$ on an open dense subset without being constant. 
This of course annihilate any hope of computing \textsc{gcd}s of 
polynomials when only approximations of them are known. Fortunately, 
there exists a standard way to recover continuity in this context: it 
consists in replacing \textsc{gcd}s by the so-called subresultants 
(whose definitions are main properties are quickly recalled below) which 
are playing an analoguous role. For this reason, in what follows, we 
will exclusively consider the problem of computing subresultants.

\noindent
\textbf{Notations.} We refer to \cite{} for a nice and rather complete 
introduction to resultants and subresultants and seek to fix notations
in this paragraph. For now, we work over a 
general ring $\ring$. We denote by $\ring_{<n}[X]$ (resp. $\ring_{\leq 
n}[X]$) the set of polynomials of degree less than $n$ (resp. less than 
or equal to $n$) with coefficients in $\ring$. Both $\ring_{<n}[X]$ and 
$\ring_{\leq n}[X]$ are free $\ring$-modules of finite rank.

@@

\begin{theo}
We assume that $\ring$ is a field. Let $A$ and $B$ be two polynomials
with coefficients in $\ring$. Let $j$ be the smallest integer such
that $R_j(A,B)$ does not vanish. Then $R_j(A,B)$ is a \textsc{gcd}
of $A$ and $B$.
\end{theo}

Since they are defined as determinants, subresultants behave well with 
respect to base change: if $f : \ring \to \ring'$ is a morphism of rings 
and $A$ and $B$ are polynomials over $\ring$ then:
$$\Res_j^{d_A,d_B}(f(A),f(B)) = f\big(\Res_j^{d_A,d_B}(A,B)\big)$$
where $f(A)$ and $f(B)$ denotes the polynomials deduced from $A$
and $B$ respectively by applying $f$ coefficient-wise. This property
is sometimes referred to as the \emph{functoriality} of subresultants.

@@

This reinterpretation is based on the so-called \emph{subresultant 
pseudo-remainder sequence} which is defined as follows. We pick $A$ and 
$B$ as above and assume that $d_A = \deg A$. Denoting by $(P \,\%\, Q)$ 
the remainder in the Euclidean division of $P$ by $Q$, we define two 
recursive sequences $(S_i)$ and $(c_i)$ as follows:
$$\left\{\begin{array}{ll}
S_0 = A, \, S_1 = B, \, c_0 = 1 \smallskip \\
\displaystyle S_{i+1} = (-s_i)^{\delta_i + 1}
s_{i-1}^{-1} \: c_i^{-\delta_i} \cdot (S_{i-1} \,\%\, S_i)
& \text{for } i \geq 1 \smallskip \\
\displaystyle c_{i+1} = s_{i+1}^{\delta_{i+1}} \cdot c_i^{1-\delta_{i+1}}
& \text{for } i \geq 0. 
\end{array}\right.$$
Here $n_i = \deg S_i$ and $s_i$ is the leading coefficient of 
$S_i$ if $i > 0$ and $s_0 = 1$ by convention. These sequences are
finite and the above recurrence applies until $S_i$ has reached the
value $0$.

% \delta_i = n_{i-1} - n_i

\begin{prop}
\label{prop:prem}
We keep the above notations. For all positive integer $i$ such that
$S_i$ is defined, we have:
$$S_i = \Res_{\deg S_{i-1} - 1} (A,B).$$
\end{prop}

\begin{rem}
Mimicing ideas behind extended Euclide's algorithm, one can define the 
``extended subresultant pseudo-remainder sequence'' as well and compute
this way at the same time the cofactors.
\end{rem}

Important simplifications occur in the common case where all scalar 
subresultants do not vanish. Under this additional assumption, one can 
prove that the degrees of the $S_i$'s decrease by one at each step, 
meaning that we $\deg S_i = d_B + 1 - i$ for all $i$. The sequence 
$(S_i)$ then stops at $i = d_B + 1$. Moreover, the $\delta_i$'s and the 
$c_i$'s are no longer necessary since we have $\delta_i = 1$ and $c_i = 
s_i$ for all $i$. The recurrence formula then becomes:
$$\left\{\begin{array}{ll}
S_0 = A, \, S_1 = B \smallskip \\
\displaystyle S_{i+1} = s_i^2 \cdot s_{i-1}^{-2}
\cdot (S_{i-1} \,\%\, S_i) & \text{for } i \geq 1.
\end{array}\right.$$
To conclude with, we remark that Proposition \ref{prop:prem} simply 
states that $R_j = S_{d_B+1-j}$; we then get this way all subresultants.

\section{Euclide-like algorithms\\are unstable}

@@

In this subsection, we estimate the loss of precision if we compute the 
$R_j$'s using the recurrence formula given by Proposition \ref{prop:prem}. 
In what follows, we are going use a \emph{flat precision model}: this 
means that a polynomial $P(X)$ is represented as:
$$P(X) = \sum_{i=1}^n a_i X^i + O(p^N)
\quad \text{with } a_i \in \Qp \text{ and } N \in \Z.$$
In other words, we shall assume that the software we are using does not
carry a precision data on each coefficient but only a unique precision
data for the whole polynomial. Concretely this means that, after having
computing a polynomial, the software truncates the precision on each
coefficients to the smallest precision. 
One can argue that this assumption is too strong (compared to usual
implementations of $p$-adic numbers). Nevertheless, it defines a 
simplified framework where computations can be performed and experiments 
show that it rather well reflects the behaviour of the loss of precision 
in Euclide-like algorithms.

Let $V_j$ be the valuation of the leading coefficient of $R_j$ and $W_j$ 
be the smallest valuation of a coefficient of $R_j$. We of course have 
$V_j \geq W_j$ and we set $\delta_j = V_j - W_j$.

\begin{prop}
\label{prop:precEuclide}
If $A$ and $B$ are known at precision $O(p^N)$ then, in the flat 
precision model, the subresultant pseudo-remainder sequence computes 
$R_j$ at precision $O(p^{N_j})$ with
$$N_j \leq N + V_{j+1} - 2 \sum_{k=j+1}^d \delta_j.$$
\end{prop}

\begin{proof}
\end{proof}

\begin{rem}
\end{rem}

@@

Proposition \ref{prop:precEuclide} gives an estimation of the loss
of precision in Euclide-like algorithms in terms of the quantities
$V_j$ and $\delta_j$. It is nevertheless \emph{a priori} not clear
how large these numbers are. The aim of this paragraph is to compute
their order of magnitude when $A$ and $B$ are picked randomly among
the set of monic polynomials of degree $d$ with coefficients in $\A$.

@@

\begin{theo}
\label{th:lawVj}
We fix $j \in \{0, \ldots, d-1\}$.
Let $X_0, \ldots, X_{d-1}$ be $d$ pairwise independant discrete random 
variables with geometric law of parameter $(1 - p^{-1})$, \emph{i.e.}
$$\P[X_i = k] = (1 - p^{-1}) \cdot p^{-k} \quad 
\text{(with } 0 \leq i < d \text{ and } k \in \N \text{)}.$$
Then $V_j$ is distributed as the random variable
$$Y_j = \sum_{i=0}^d \min(X_{j-i}, X_{j-i+1}, \ldots, X_{j+i})$$
where, by convention, $X_i = +\infty$ if $i < 0$ and $X_i = 0$ if 
$i \geq d$.
\end{theo}

\begin{rem}
The above Theorem does not say anything about the correlations between
the $X_j$'s. In particular, we emphasize that it is \emph{false} that
the tuple $(V_{d-1}, \ldots, V_0)$ is distributed as $(Y_{d-1}, \ldots,
Y_0)$. For instance, an explicit computation shows that $(V_{d-1}, 
V_{d-2})$ is distributed as 
$$(X, \, Y + \min(Y, [X/2]))$$
where $X$ and $Y$ are two independant discrete random variables with 
geometric law of parameter $(1 - p^{-1})$ and the notation $[\cdot]$ 
stands for the integer part function. In particular, we observe that 
the couple $(2,1)$ does not occur almost surely\footnote{Actually, one
can prove that it never occurs.} although the events $\{V_{d-1} = 2\}$ 
and $\{V_{d-2} = 1\}$ both occur with positive probability.
\end{rem}

>>>>

\usepackage{amsmath,amssymb}
%\usepackage{amsrefs}
\usepackage[usenames,dvipsnames]{color}
\usepackage{stmaryrd}
\usepackage{enumerate}
%\usepackage[algoruled,vlined,english,linesnumbered]{algorithm2e}
\usepackage[pdfpagelabels,colorlinks=true,citecolor=blue]{hyperref}
\usepackage{comment}
\usepackage{tikz}
\usepackage{bbm}
\usepackage{multirow}
\renewcommand*{\theHsection}{\thesection}

@@

\newcommand{\ring}{\mathfrak A}
\newcommand{\fracring}{\mathfrak K}
\renewcommand{\prec}{\text{\rm prec}}
\renewcommand{\val}{\text{\rm val}}
\newcommand{\id}{\text{\rm id}}
\newcommand{\Res}{\text{\rm Res}}
\newcommand{\lc}{\text{\rm lc}}
\newcommand{\Card}{\text{\rm Card}\:}

@@

\title{Resultants and subresultants of p-adic polynomials}

@@

\begin{abstract}
We address the problem of the stability of the computations of resultants 
and subresultants of polynomials defined over complete discrete valuation 
rings (\emph{e.g.} $\Zp$ or $k[[t]]$ where $k$ is a field). We prove that 
Euclide-like algorithms are highly unstable in average and we explain, in
many cases, how one can stabilize them without sacrifying the complexity.
On the way, we completely determine the distribution of the valuation of
the subresultants of two random monic $p$-adic polynomials having the
same degree.
\end{abstract}

@@

 \vspace{1mm}
 \noindent
 {\bf Keywords:} $p$-adic precision, subresultants
\medskip

@@

As wonderfully illustrated by the success of Kedlaya-type counting points 
algorithms \cite{kedlaya}, $p$-adic technics are gaining nowadays more and more 
popularity in computer science, and more specifically in symbolic 
computation. A crucial issue when dealing with $p$-adics is those of 
stability. Indeed, just like real numbers, $p$-adic numbers are by nature 
infinite and thus need to be truncated in order to fit in the memory of a 
computer. The level of truncation is called the \emph{precision}. Usual 
softwares implementing $p$-adics (\emph{e.g.} \textsc{magma} \cite{magma}, 
\textsc{pari} \cite{pari}, \textsc{sage} \cite{sage}) 
generally tracks the precision as follows: an individual precision is 
attached to any $p$-adic variable and this precision is updated after 
each basic arithmetic operation (addition and multiplication). This way 
of tracking precision can be seen as the analogue of the arithmetic 
intervals in the real setting.

In the paper \cite{padicprec}, the authors propose a new framework to control 
$p$-adic precision. The aim of this paper is to illustrate the technics 
of \emph{loc. cit.} on the concrete example of computation of 
\textsc{gcd}s and subresultants of $p$-adic polynomials. There is 
actually a real need to do this due to the combination of two reasons: on 
the one hand, computating \textsc{gcd}s is a very basic operation for 
which it cannot be acceptable to have important instability whereas, on 
the other hand, easy experimentations show that all standard algorithms 
for this task (\emph{e.g.} extended Euclide's algorithm) are \emph{very} 
unstable.
%
\begin{figure}
\renewcommand{\arraystretch}{1.2}
\begin{center}
\begin{tabular}{|c||c|c|}
\hline
\multirow{2}{*}{\hspace{0.2cm}Degree\hspace{0.2cm}\null} & 
\multicolumn{2}{c|}{\hspace{0.3cm}Loss of precision (in number of digits)\hspace{0.3cm}\null}\\
\cline{2-3}
& \hspace{0.3cm}Euclide algorithm\hspace{0.3cm}\null & expected \\
\hline
$5$ & $\phantom{00}6.3$ & $3.1$ \\
$10$ & $\phantom{0}14.3$ & $3.2$ \\
$25$ & $\phantom{0}38.9$ & $3.2$ \\
$50$ & $\phantom{0}79.9$ & $3.2$ \\
$100$ & $160.0$ & $3.2$ \\
\hline
\end{tabular}
\end{center}

\vspace{-0.3cm}

\caption{Average loss of precision when computing the {\sc gcd}
of two random monic polynomial of fixed degree over $\Z_2$.}
\label{fig:precision}

\vspace{-0.3cm}
\end{figure}
%
Figure \ref{fig:precision} illustrates the instability of extended Euclide's 
algorithm when it is called on random inputs which are monic $2$-adic 
polynomials of fixed degree. Looking at the last line, we see that 
extended Euclide's algorithm outputs the B\'ezout coefficients of two 
monic $2$-adic polynomials of degree $100$ with an average loss of 
$161.1$ significant digits by coefficient whereas a stable 
algorithm should only loose \todo{$4$} digits in average. This 
``theoretical'' loss is computed as the double of the valuation of the 
resultant. Indeed Cramer-like formulae imply that B\'ezout coefficients 
can be computed by performing a unique division by the resultant, 
inducing then only the aforementioned loss of precision. Examining the 
table a bit more, we furthermore observe that the ``practical'' loss of 
precision due to Euclide's algorithm seems to grow linearly with respect 
to the degree of the input polynomials whereas the ``theoretical'' loss 
seems to be independant of it. In other words, the unstability of 
Euclide's algorithm is becoming more and more critical when the degree 
of the input increases.

@@

\noindent
{\bf Content of the paper.}
The aim of this article is twofold. We first provide in \S 
\ref{sec:unstable} a theoretical study of the instability phenomenon 
described above and give strong evidences that the loss of precision 
grows linearly with respect to the degree of the input polynomials, as we 
observed empirically. In doing so, we determine the distribution of the 
valuation of the subresultants of random monic polynomials over $\Zp$ 
(\emph{cf} Theorem \ref{th:lawVj}). This is an independant result which 
has its own interest.

Our second goal, which is carried out in \S \ref{sec:stable}, is to rub 
out these unexpected losses of precision. Making slight changes to the 
standard subresultants pseudo-remainder sequence algorithm and using in 
an essential way the results of \cite{padicprec}, we manage to design a stable 
algorithm for computing all subresultants of two monic polynomials over 
$\Zp$ (satisfying an additional mild assumption). This basically allows 
to stably compute \textsc{gcd}s assuming that the degree of the 
\textsc{gcd} is known in advance.

@@

The aim of this section is to introduce the setting we shall work in 
throughout this paper (which is a bit more general than those considered 
in the introduction).

@@

\noindent
\textbf{Definitions.}
Let $\A$ be a \emph{discrete valuation ring} (DVR for short), that is a 
domain equipped with a map $\val : \A \to \Z \cup \{+\infty\}$ --- the 
so-called \emph{valuation} --- satisfying the four axioms: (1)~$\val(x) 
= +\infty$ iff $x = 0$, (2)~$\val(xy) = \val(x) + \val(y)$ and 
(3)~$\val(x+y) \geq \min(\val(x), \val(y))$, (4)~any element of 
valuation $0$ is invertible. In the sequel, we shall assume that the
valuation is normalized so that it takes the value $1$. We recall that a 
DVR admits a unique maximal ideal $\m$, consisting of elements of 
positive valuation. Furthermore, this ideal is principal and generated 
by any element of valuation $1$. Such an element is called a 
\emph{uniformizer}. Let us fix one of them and denote it by $\pi$. The 
\emph{residue field} of $W$ is the quotient $W/\m = W/\pi W$ and we 
shall denote it by $k$.

The valuation defines a distance $d$ on $\A$ by letting $d(x,y) = 
e^{-\val(x-y)}$ for all $x, y \in \A$. We say that $\A$ is \emph{complete}
if it is complete with respect to the previous distance. Assuming that
$\A$ is complete, any element $x \in \A$ can be written uniquely as a 
convergent series:
\begin{equation}
\label{eq:expandCDVR}
x = x_0 + x_1 \pi + x_2 \pi^2 + \cdots + x_n \pi^n + \cdots
\end{equation}
where the $x_i$'s lie in a fixed set of representatives of classes
modulo $\pi$. Therefore, as an additive group, $W$ is isomorphic to 
the set of sequences $\N \to k$. On the contrary, the multiplicative
structure may vary.
Let $K$ denote the fraction field of $\A$. The valuation $v$ extends 
uniquely to $K$ by letting $\val(\frac x y) = \val(x) - \val(y)$. 
Moreover, it follows from the axiom~(4) that $K$ is obtain from $\A$ by 
inverting $\pi$. Thus, any element of $K$ can be uniquely written as an 
infinite sum $x = \sum_{i=v}^\infty x_i \pi^i$ where $v$ is some 
relative integer and the $x_i$'s are as above. The valuation of $x$ can 
be easily read on this writing: it is the smallest integer $v$ such that 
$x_i \not\equiv 0 \pmod \pi$.

The two main examples of complete discrete valuation rings are the 
ring of $p$-adic integers $\Zp$ and the ring of formal series 
$k[[x]]$ over a field $k$. Their fraction fields are $\Q_p$ and
$k((x))$ respectively.

@@

\noindent
\textbf{Precision.}
The memory of a computer being necessarily finite, it cannot be possible 
to find a way to represent exhaustively all elements of $\A$. 
Very often, mimicing what we do for real numbers, we choose to truncate 
the expansion \eqref{eq:expandCDVR} at some finite level. Concretely,
this means that we work with approximations of elements of $\A$ of the
form
$\sum_{i=0}^{N-1} x_i \pi^i + O(\pi^N)$
where $N$ is some integer called the \emph{precision} of $x$ and 
the term $O(\pi^N)$ means that the $x_i$'s with $i \geq N$ are not 
specified. Hence, the notation above does not represent a unique
element of $\A$ but all elements lying in the ball having center
$\sum_{i=0}^{N-1} x_i \pi^i$ and radius $e^{-N v(\pi)}$. In other
words, on a computer, we do not work with elements of $\A$ but with
balls in $\A$ which are much more tractable.

\subsection{Subresultants}
\label{subsec:subres}

A first issue when dealing with numerical computations of \textsc{gcd}s 
of polynomials over $\A$ is that the \textsc{gcd} function is not 
continuous: it takes the value $1$ on an open dense subset without being 
constant. This of course annihilate any hope of computing \textsc{gcd}s 
of polynomials when only approximations of them are known. Fortunately, 
there exists a standard way to recover continuity in this context: it 
consists in replacing \textsc{gcd}s by the so-called subresultants which 
are playing an analoguous role. For this reason, in what follows, we will 
exclusively consider the problem of computing subresultants.

@@

\noindent
\textbf{Notations.} We refer to \cite{resultant} for a nice and rather complete 
introduction to resultants and subresultants and seek to fix notations
in this paragraph. For now, we work over a 
general ring $\ring$. We denote by $\ring_{<n}[X]$ (resp. $\ring_{\leq 
n}[X]$) the set of polynomials of degree less than $n$ (resp. less than 
or equal to $n$) with coefficients in $\ring$. Both $\ring_{<n}[X]$ and 
$\ring_{\leq n}[X]$ are free $\ring$-modules of finite rank.

@@

\begin{theo}
We assume that $\ring$ is a field. Let $A$ and $B$ be two polynomials
with coefficients in $\ring$. Let $j$ be the smallest integer such
that $\Res_j(A,B)$ does not vanish. Then $\Res_j(A,B)$ is a \textsc{gcd}
of $A$ and $B$.
\end{theo}

Since they are defined as determinants, subresultants behave well with 
respect to base change: if $f : \ring \to \ring'$ is a morphism of rings 
and $A$ and $B$ are polynomials over $\ring$ then
$\Res_j^{d_A,d_B}(f(A),f(B)) = f\big(\Res_j^{d_A,d_B}(A,B)\big)$
where $f(A)$ and $f(B)$ denotes the polynomials deduced from $A$
and $B$ respectively by applying $f$ coefficient-wise. This property
is sometimes referred to as the \emph{functoriality} of subresultants.

@@

This reinterpretation is based on the so-called \emph{subresultant 
pseudo-remainder sequence} which is defined as follows. We pick $A$ and 
$B$ as above and assume that $d_A = \deg A$. Denoting by $(P \,\%\, Q)$ 
the remainder in the Euclidean division of $P$ by $Q$, we define two 
recursive sequences $(S_i)$ and $(c_i)$ as follows:
$$\left\{\begin{array}{ll}
S_{-1} = A, \, S_0 = B, \, c_{-1} = 1 \smallskip \\
\displaystyle S_{i+1} = (-s_i)^{\delta_i + 1}
s_{i-1}^{-1} \: c_i^{-\delta_i} \cdot (S_{i-1} \,\%\, S_i)
& \text{for } i \geq 1 \smallskip \\
\displaystyle c_{i+1} = s_{i+1}^{\delta_{i+1}} \cdot c_i^{1-\delta_{i+1}}
& \text{for } i \geq 0. 
\end{array}\right.$$
Here $n_i = \deg S_i$ and $s_i$ is the leading coefficient of 
$S_i$ if $i \geq 0$ and $s_{-1} = 1$ by convention. These sequences 
are finite and the above recurrence applies until $S_i$ has reached 
the value $0$.

\begin{prop}
\label{prop:prem}
With the above notations, we have:
$$\begin{array}{rcll}
\Res_j(A,B) & = & S_i & \text{if } j = n_{i-1} - 1 \smallskip \\
& = & 0 & \text{if } n_i < j < n_{i-1} - 1 < j \\
& = & \!\big(\frac{s_i}{s_{i-1}}\big)^{\delta_i-1} \cdot S_i & 
\text{if } j = n_i
\end{array}$$
for all $i$ such that $S_i$ is defined.
\end{prop}

\begin{rem}
The Proposition \ref{prop:prem} computes \emph{all} subresultants. We
note moreover that, in the common case where $n_{i-1} = n_i - 1$, the
two formulas giving $\Res_{n_i}(A,B)$ agree.

Mimicing ideas behind extended Euclide's algorithm, one can define the 
``extended subresultant pseudo-remainder sequence'' as well and compute
this way at the same time the cofactors.
\end{rem}

Important simplifications occur in the common case where all scalar 
subresultants do not vanish. Under this additional assumption, one can 
prove that the degrees of the $S_i$'s decrease by one at each step, 
meaning that we $\deg S_i = d_B - i$ for all $i$. The sequence 
$(S_i)$ then stops at $i = d_B$. Moreover, the $\delta_i$'s and the 
$c_i$'s are no longer necessary since we have $\delta_i = 1$ and $c_i
= s_i$ for all $i$. The recurrence formula then becomes:
$$S_{i+1} = s_i^2 \cdot s_{i-1}^{-2}
\cdot (S_{i-1} \,\%\, S_i) \quad \text{for } i \geq 1.$$
To conclude with, we remark that Proposition \ref{prop:prem} simply 
states that $R_j = S_{d_B-j}$.

\section{Euclide-like algorithms\\are unstable}
\label{sec:unstable}

@@

In this subsection, we estimate the loss of precision if we compute the 
$R_j$'s using the recurrence formula given by Proposition \ref{prop:prem}. 
In what follows, we are going use a \emph{flat precision model}: this 
means that a polynomial $P(X)$ is represented as:
$$P(X) = \sum_{i=1}^n a_i X^i + O(p^N)
\quad \text{with } a_i \in \Qp \text{ and } N \in \Z.$$
In other words, we assume that the software we are using does not
carry a precision data on each coefficient but only a unique precision
data for the whole polynomial. Concretely this means that, after having
computing a polynomial, the software truncates the precision on each
coefficients to the smallest precision. 
One can argue that this assumption is too strong (compared to usual
implementations of $p$-adic numbers). Nevertheless, it defines a 
simplified framework where computations can be performed and experiments 
show that it rather well reflects the behaviour of the loss of precision 
in Euclide-like algorithms.

Let $V_j$ be the valuation of the scalar $j$-th subresultant of $(A,B$) 
and $W_j$ be the smallest valuation of a coefficient of $R_j$. We of 
course have $V_j \geq W_j$ and we set $\delta_j = V_j - W_j$.

\begin{prop}
\label{prop:precEuclide}
Let $A$ and $B$ as above. We assume that all scalar subresultants of 
$(A,B)$ do not vanish. Then, in the flat precision model, the 
subresultant pseudo-remainder sequence computes $R_j$ at precision 
$O(p^{N_j})$ with:
$$N_j \leq N + V_{j+1} - 2 \cdot (\delta_{j+1} + \delta_{j+2} + \cdots
+ \delta_{d-1}).$$
\end{prop}

\begin{proof}
By our assumptions, the subresultant pseudo-remainder sequence 
is defined by the recurrence:
$$R_{j-1} = r_j^2 \cdot r_{j+1}^{-2} \cdot 
(R_{j+1} \,\%\, R_j).$$
Using that $R_j$ and $R_{j-1}$ have the expected degrees, the
remainder $(R_{j+1} \,\%\, R_j)$ is computed as follows:
$$\begin{array}{rrcl}
\text{we set:} & S & = & R_{j+1} - r_{j+1} \cdot r_j^{-1} \cdot R_j \smallskip \\
\text{and we have:} & R_{j+1} \,\%\, R_j & = & S - s \cdot r_j^{-1} \cdot R_j
\end{array}$$
where $s$ is the coefficient of degree $j$ of $S$. Let us first estimate 
the precision of $S$. Recall for that the relative precision is defined 
as the difference between the (absolute) precision and the valuation: 
intuitively, it is the number of significant digits. It is easy to check 
that the relative precision of a product is the minimum of those of the 
factors. 
Thus the relation precision of $r_{j+1} \cdot r_j^{-1} \cdot R_j$ is
$\min(N_{j+1} - V_{j+1}, N_j - V_j)$. Its absolute precision is then
$M = \min(N_{j+1} - \delta_j, N_j - \delta_j + V_{j+1} - V_j)$. This
quantity is also the precision of $S$ since the other summand
$R_{j+1}$ is known with higher precision. Repeating the argument, we
find that the precision of $(R_{j+1} \,\%\, R_j)$ is equal to 
$\min(M - \delta_j, N_j - \delta_j + \val(s) - V_j)$ and therefore is 
lower bounded by $M - \delta_j \leq N_j - 2 \delta_j + V_{j+1} - V_j$.
We derive for this $N_{j-1} \leq N_j - 2 \delta_j - V_{j+1} + V_j$ and
the proposition finally follows by summing up these inequalities.
\end{proof}

The difference 
$N - N_0 = - V_1 + 2 \sum_{k=1}^d \delta_j$
is a lower bound on the number of digits lost after having computed the 
resultant using the subresultant pseudo-remainder sequence algorithm. In 
the next subsection (\emph{cf} Corollary \ref{cor:Vj}), we shall see that 
$V_1$ and all $\delta_j$'s are 
approximatively equal to $\frac 1 {p-1}$ in average. The loss of 
precision then grows linearly with respect to $d$. This confirms the 
precision benchmarks shown in Figure \ref{fig:precision}.
We emphasize one more time that this loss of precision is \emph{not}
intrinsic but an artefact of the algorithm we have used; indeed, one
should not loose any precision when computing resultants because they
are given by polynomial expressions.

@@

Proposition \ref{prop:precEuclide} gives an estimation of the loss
of precision in Euclide-like algorithms in terms of the quantities
$V_j$ and $\delta_j$. It is nevertheless \emph{a priori} not clear
how large these numbers are. The aim of this paragraph is to compute
their order of magnitude when $A$ and $B$ are picked randomly among
the set of monic polynomials of degree $d$ with coefficients in $\A$.
In what follows, we assume that the residue field of $\A$ is finite
and we use the letter $q$ to denote its cardinality.

@@

\begin{theo}
\label{th:lawVj}
We fix $j \in \{0, \ldots, d-1\}$.
Let $X_0, \ldots, X_{d-1}$ be $d$ pairwise independant discrete random 
variables with geometric law of parameter $(1 - q^{-1})$, \emph{i.e.}
$$\P[X_i = k] = (1 - q^{-1}) \cdot q^{-k} \quad 
\text{(with } 0 \leq i < d \text{ and } k \in \N \text{)}.$$
Then $V_j$ is distributed as the random variable
\begin{center}
$\displaystyle Y_j = \sum_{i=0}^d \min(X_{j-i}, X_{j-i+1}, \ldots, X_{j+i})$
\end{center}
\noindent
with $X_i = +\infty$ if $i < 0$ and $X_i = 0$ if $i \geq d$.
\end{theo}

\begin{rem}
The above Theorem does not say anything about the correlations between
the $X_j$'s. In particular, we emphasize that it is \emph{false} that
the tuple $(V_{d-1}, \ldots, V_0)$ is distributed as $(Y_{d-1}, \ldots,
Y_0)$. For instance, one can prove that $(V_{d-1}, 
V_{d-2})$ is distributed as 
$(X, \, X' + \min(X', [X/2]))$
where $X$ and $X'$ are two independant discrete random variables with 
geometric law of parameter $(1 - q^{-1})$ and the notation $[\cdot]$ 
stands for the integer part function. In particular, we observe that 
the couple $(2,1)$ does not occur almost surely although the events 
$\{V_{d-1} = 2\}$ and $\{V_{d-2} = 1\}$ both occur with positive 
probability.

Nonetheless, a consequence of Proposition \ref{prop:distribk} below is 
that the variables $\bar V_j = \mathbbm{1}_{\{V_j = 0\}}$ are mutually 
independant.
\end{rem}
